---
title: A Statistical Perspective of Non-Inferiority Trials
author:
  - name: Áine Glynn
    department: School of Mathematical and Statistical Sciences
    affiliation: University of Galway
  - name: Filip Kłosowski
    department: School of Mathematical and Statistical Sciences
    affiliation: University og Galway
abstract: |
  Single-arm performance goal studies are widely used in medical device evaluation when a randomised active-control trial is impractical or ethically unjustifiable. In this setting, the trial conclusion depends on whether a confidence interval for the observed event proportion satisfies a pre-specified benchmark derived from historical evidence. Despite their apparent simplicity, the conclusions of such studies are sensitive to three interconnected methodological choices: the value of the performance goal, the confidence interval method used to construct the decision rule, and the sample size calculation underlying the study design. This thesis examines each of these choices in the context of a retrospective case study of the BioMimics 3D vascular stent, a helical nitinol stent developed for the treatment of femoropopliteal peripheral artery disease.Performance goals for a 30-day composite safety endpoint and a 12-month Rutherford classification efficacy endpoint were derived through a generalised linear mixed model meta-analysis of three historical femoropopliteal bare nitinol stent studies comprising 116 patients in total. Using confidence interval boundaries rather than point estimates to account for uncertainty in the historical evidence yielded a maximum acceptable 30-day composite adverse event rate of 12% and a minimum acceptable 12-month Rutherford success proportion of 88%. The limited evidence base produced wide uncertainty around both estimates, illustrating that performance goals carry inherent uncertainty that is not always visible when they are reported as fixed values.
  A simulation study evaluated six confidence interval methods for binary proportions across a grid of true event proportions and sample sizes relevant to the case study endpoints. The Wald interval was the only method to consistently fall below the nominal 95% coverage level and was judged unsuitable as a decision tool in this setting. Among the remaining methods, Wilson Score and Jeffreys achieved coverage closest to the nominal level with the narrowest intervals, while Clopper-Pearson, Agresti-Coull and the continuity-corrected Wilson Score were conservative with progressively wider intervals. These differences in coverage behaviour translated into meaningful differences in sample size requirements. Under an approved performance goal of 11%, an assumed true complication proportion of 5% and a target power of 90%, Clopper-Pearson was used as the primary method, requiring 209 evaluable patients at 89.7% achieved power. Among methods meeting the nominal coverage level, the required sample size ranged from 198 patients under Jeffreys to 247 under the continuity-corrected Wilson Score.
  A Bayesian historical borrowing analysis demonstrated that incorporating prior evidence through a Beta(2, 48) prior reduced the required sample size from 209 to 79 patients, a reduction of approximately 62%. This reduction was sensitive to the weight assigned to the prior, with stronger borrowing increasing dependence on the exchangeability assumption and the risk of inflating the Type I error rate when historical and current populations are not sufficiently comparable.
  An interactive R Shiny application, PG-Power, was developed to implement the sample size framework and make the sensitivity of required sample size to key design assumptions accessible and reproducible. The application supports confidence interval method comparison, power curve visualisation, sensitivity analysis and report generation for single-arm performance goal studies.
  The findings demonstrate that the performance goal value, confidence interval method and sample size procedure are not independent design choices and that each can alter the conclusion of a performance goal study independently of the underlying clinical data. Transparent derivation and pre-specification of these choices in the SAP is essential to the credibility and reproducibility of performance goal studies in medical device evaluation.

keywords:
  - binomial confidence intervals
  - performance goals
  - single-arm clinical trials
  - medical device evaluation
  - sample size calculation
  - Bayesian historical borrowing
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
---

# 1. Introduction

## 1.1 Non-inferiority Trials

Non-inferiority trials arise when an effective treatment or device already exists and a placebo-controlled comparison would be unethical, impractical or clinically uninformative(1–3). Rather than demonstrating superiority, the goal is to show that any loss of efficacy associated with the new intervention remains within a clinically acceptable limit(1–3). This limit, known as the non-inferiority margin, represents the largest reduction in efficacy considered tolerable in exchange for other benefits such as improved safety, simpler delivery, lower cost or better patient adherence(1,2,4).
The trial conclusion depends on both the observed treatment effect and the uncertainty surrounding it(1,2,5). In practice, this is assessed by comparing a confidence interval for the treatment effect against the pre-specified margin(1,2,6). For favourable endpoints, where higher values indicate better performance, the lower confidence bound is the relevant quantity, while for unfavourable endpoints the upper bound is used(1,2). This one-sided structure distinguishes non-inferiority trials from superiority and equivalence designs, and places the choice of margin and confidence interval method at the centre of the trial's interpretation(1–3,6).

## 1.2	Performance Goals in Medical Device Studies

Performance goal studies are closely related to non-inferiority designs and are particularly common in medical device evaluation(1,7,8). Rather than comparing the new intervention against an active control, the observed outcome is compared with a pre-specified benchmark derived from historical evidence and clinical judgement(7,9). This benchmark defines the minimum level of performance considered acceptable for the device under evaluation(7,9).
The performance goal approach is most useful when a randomised active-control trial is difficult to justify or conduct(7,8). However, the absence of a concurrent comparator makes the choice of benchmark especially consequential(7,10). Historical studies may differ from the current trial in ways that affect the validity of the comparison, including differences in the target population, endpoint definitions, follow-up duration and broader clinical practice(7,10). These differences must be carefully considered when constructing and justifying a performance goal.
The direction of the performance goal depends on the nature of the endpoint(7,9). For efficacy endpoints, where higher event proportions reflect better performance, the estimated proportion and its confidence interval must exceed a minimum acceptable level(7,9). For safety endpoints, where lower proportions are desirable, the estimated proportion and its confidence interval must remain below a maximum acceptable level of harm(7,9). In either case, a performance goal should not be treated as a simple numerical cut-off(5,7,9). It represents a clinical and statistical threshold whose validity depends on the endpoint being measured, the quality and consistency of the historical evidence, and the level of uncertainty considered acceptable for regulatory decision-making(7,8,10).

## 1.3 Motivation for Study

Although non-inferiority and performance goal designs appear straightforward in principle, their conclusions can be sensitive to several methodological choices(1–3). Small variations in any of these decisions can alter the interpretation of the same clinical data and, consequently, the study conclusion(1,2).
This sensitivity is particularly consequential in medical device studies, where single-arm performance goal designs are common and the absence of a concurrent comparator means the historical benchmark carries much of the evidential weight(7,8,11). If the benchmark is too lenient, a device may appear acceptable despite a clinically meaningful loss of performance. If it is too strict, a device that performs well may require an impractically large sample size or fail to meet the criterion regardless(4,7,10). In either case, the benchmark itself shapes the conclusion before a single patient is enrolled.
Previous reviews have raised concerns about how consistently these choices are reported. Piaggio et al. found that only around one-fifth of 332 non-inferiority and equivalence trials published between 1990 and 2000 provided a suitable rationale for the chosen margin(6). A later review of 232 drug trials found that while nearly all reports specified a margin, only 24% explained how it had been determined(6). Together these findings point to a recurring problem: margins and performance goals are often presented as settled values without making the reasoning behind them transparent(2,6).
This study is motivated by that problem. Using the BioMimics 3D vascular stent trial as a case study, it examines how performance goal selection, confidence interval choice and sample size calculation interact to influence conclusions in a single-arm medical device setting(1,5–7).

## 1.4 Case Study: Biomimics 3D Vascular Stent

The BioMimics 3D vascular stent trial serves as the primary case study throughout this analysis. Rather than attempting to reproduce the original trial, the case study is used as a real medical device setting in which to examine how statistical choices surrounding performance goals influence the interpretation of device performance.
Clinical and regulatory context for the analysis was obtained through meetings with a lead scientist involved in the original trial. These discussions clarified how the device was intended to be evaluated in practice and highlighted that the statistical design could not be reduced to a purely numerical exercise- it had to reflect clinical judgement, regulatory expectations and practical feasibility alongside the available evidence.

## 1.5 Study Aims

This study has five objectives:

1.	To derive evidence-based performance goals for a single-arm medical device study through meta-analysis of comparable historical device studies;
2.	To compare confidence interval methods for binary endpoints in terms of coverage probability and interval width;
3.	To evaluate how performance goal selection and confidence interval choice affect sample size requirements and non-inferiority conclusions;
4.	To develop an interactive Shiny application, PG-Power, to support sample size planning and confidence interval method selection in performance goal studies;
5.	To explore the potential role of Bayesian historical borrowing as a means of reducing the required sample size.

# 2.	Background and Statistical Framework

## 2.1 Superiority, Equivalence and Non-Inferiority

Superiority, equivalence and non-inferiority trials address distinct clinical questions and should not be interpreted interchangeably(12,13). Let p_Tdenote the success probability in the investigational treatment group and p_Cdenote the success probability in the comparator group. For a binary success endpoint, the treatment effect is defined as(13,14):

$$\theta = p_T - p_C$$

A superiority trial evaluates whether the investigational treatment performs better than the comparator. The null and alternative hypotheses are(1,12,13):

$$H_0: \theta \leq 0 \qquad H_1: \theta > 0$$
Superiority is concluded when the lower confidence bound for θ lies above zero(1).

An equivalence trial asks whether the difference between treatments is small enough to be considered clinically unimportant in either direction. Denoting the equivalence margin as $$Δ_E$$, the hypotheses are(6,12,13)

$$H_0: \theta \leq -\Delta_E \text{ or } \theta \geq \Delta_E \qquad H_1: -\Delta_E < \theta < \Delta_E$$

Equivalence is concluded when the confidence interval for θ lies entirely within the equivalence margins(6,12,13):

$$-\Delta_E < \text{CI}_{1-\alpha}(\theta) < \Delta_E$$

A non-inferiority trial is less restrictive, focusing only on differences in the unfavourable direction. Denoting the non-inferiority margin as Δ_NI, the largest clinically acceptable loss of efficacy, the hypotheses are(1,2,12,13):

$$H_0: \theta \leq -\Delta_{NI} \qquad H_1: \theta > -\Delta_{NI}$$

Non-inferiority is concluded when the lower confidence bound lies above the margin(1,6,12,13):

$$L_{1-\alpha}(\theta) > -\Delta_{NI}$$

These three designs are often confused in practice, but the distinction matters. Failure to demonstrate superiority does not imply equivalence or non-inferiority- a non-significant superiority test may simply reflect insufficient power or imprecision rather than genuine similarity between treatments(1,2,12). Each design therefore requires its own pre-specified margin, endpoints, analysis population and decision rule, and conclusions should not be switched between frameworks after the data have been collected(5,6,13).



# Headings: first level
\label{sec:headings}

You can use directly LaTeX command or Markdown text. 

LaTeX command can be used to reference other section. See Section \ref{sec:headings}.
However, you can also use **bookdown** extensions mechanism for this.

## Headings: second level

You can use equation in blocks

$$
\xi _{ij}(t)=P(x_{t}=i,x_{t+1}=j|y,v,w;\theta)= {\frac {\alpha _{i}(t)a^{w_t}_{ij}\beta _{j}(t+1)b^{v_{t+1}}_{j}(y_{t+1})}{\sum _{i=1}^{N} \sum _{j=1}^{N} \alpha _{i}(t)a^{w_t}_{ij}\beta _{j}(t+1)b^{v_{t+1}}_{j}(y_{t+1})}}
$$

But also inline i.e $z=x+y$

### Headings: third level

Another paragraph. 


# Examples of citations, figures, tables, references
\label{sec:others}

You can insert references. Here is some text [@kour2014real; @kour2014fast] and see @hadash2018estimate.

The documentation for \verb+natbib+ may be found at

You can use custom blocks with LaTeX support from **rmarkdown** to create environment.

::: {.center latex=true}
  <http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}>
:::

Of note is the command \verb+\citet+, which produces citations
appropriate for use in inline text.  

You can insert LaTeX environment directly too.

\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}

produces

\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}


## Figures

You can insert figure using LaTeX directly. 

See Figure \ref{fig:fig1}. Here is how you add footnotes. [^Sample of the first footnote.]

\begin{figure}
  \centering
  \fbox{\rule[-.5cm]{4cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
  \label{fig:fig1}
\end{figure}

But you can also do that using R.

```{r fig2, fig.cap = "Another sample figure"}
plot(mtcars$mpg)
```

You can use **bookdown** to allow references for Tables and Figures.


## Tables

Below we can see how to use tables. 

See awesome Table~\ref{tab:table} which is written directly in LaTeX in source Rmd file.

\begin{table}
 \caption{Sample table title}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule(r){1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
  \label{tab:table}
\end{table}

You can also use R code for that.

```{r}
knitr::kable(head(mtcars), caption = "Head of mtcars table")
```


## Lists

- Item 1
- Item 2 
- Item 3
